---
title: "Differential Gene Expression Analysis for All Samples"
author: "`r params$authors`"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 3
    toc-location: left
    html-math-method: katex
    code-fold: true
    code-tools: true
    embed-resources: false
    page-layout: full
    html-table-processing: none
    theme: lumen
bibliography: references.bib
params:
  salmon_dir: ""
  samplesheet_path: ""
  organism: ""
  spikein: ""
  application: ""
  name: ""
  authors: ""
---

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
library(DESeq2)
library(dplyr)
library(ggplot2)
library(plotly)
library(viridis)
library(DT)
library(tximport)
library(data.table)
library(openxlsx)
library(ggrepel)
library(tibble)

salmon_dir <- params$salmon_dir
samplesheet <- read.csv(params$samplesheet_path, stringsAsFactors = FALSE)
rownames(samplesheet) <- samplesheet$sample
application <- params$application
organism <- params$organism
spikein <- params$spikein
norm_spikein_ercc <- isTRUE(grepl("ERCC", spikein, ignore.case = TRUE))

if (identical(application, "nfcore_RNAseq")) {
  counts_from_abundance <- "lengthScaledTPM"
  length_correction <- TRUE
} else {
  counts_from_abundance <- "no"
  length_correction <- FALSE
}

tx2gene_file <- file.path(salmon_dir, "tx2gene.tsv")

stopifnot(
  nzchar(salmon_dir),
  is.data.frame(samplesheet),
  file.exists(tx2gene_file)
)

if (organism == "hsapiens") {
  library(org.Hs.eg.db)
} else if (organism == "mmusculus") {
  library(org.Mm.eg.db)
} else if (organism == "rnorvegicus") {
  library(org.Rn.eg.db)
} else if (organism == "sscrofa" && requireNamespace("org.Ss.eg.db", quietly = TRUE)) {
  library(org.Ss.eg.db)
} else {
  warning("No annotation package available for organism: ", organism, ". Proceeding without OrgDb-dependent analyses.")
}
```

# DESeq2

This section estimates size factors and dispersions with DESeq2 and reports results from a model with intercept only (design `~ 1`), which is suitable for exploratory summaries rather than group contrasts [@love2014moderated].

```{r DESeq2, echo=TRUE, warning=FALSE, message=FALSE}
tx2gene <- fread(
  tx2gene_file,
  col.names = c("transcript_id", "gene_id", "gene_name")
)
files <- file.path(salmon_dir, samplesheet$sample, "quant.sf")
names(files) <- samplesheet$sample
if (!all(file.exists(files))) {
  stop(
    "Missing salmon quant files. ",
    "salmon_dir='", salmon_dir, "', ",
    "samplesheet_path='", samplesheet_path, "'.",
    call. = FALSE
  )
}
txi <- tximport(
  files,
  type = "salmon",
  tx2gene = tx2gene[, c(1, 2)],
  countsFromAbundance = counts_from_abundance
)

if (length_correction) {
  dds <- DESeqDataSetFromTximport(
    txi,
    colData = samplesheet,
    design = ~1
  )
} else {
  dds <- DESeqDataSetFromMatrix(round(txi$counts), samplesheet, ~1)
}

dds <- DESeq(dds)

if (norm_spikein_ercc) {
  ercc <- fread("thermofisher_LSG_manuals_cms_095046.txt")
  non_ercc <- !(rownames(dds) %in% paste0(ercc$`ERCC ID`, "_", "gene"))
  dds <- dds[non_ercc, ]
}

res <- results(dds, alpha = 0.05)
res <- as.data.frame(res)
res <- res[complete.cases(res), ]
res$gene_symbol <- tx2gene$gene_name[match(rownames(res), tx2gene$gene_id)]
res$gene_id <- rownames(res)
res$Significant <- ifelse(res$padj < 0.05, "Yes", "No")
res <- res[, c(
  "gene_id",
  "gene_symbol",
  "baseMean",
  "log2FoldChange",
  "lfcSE",
  "stat",
  "pvalue",
  "padj",
  "Significant"
)]
normalized_counts <- counts(dds, normalized = TRUE)
out_norm <- merge(
  tx2gene[, -1],
  rownames_to_column(as.data.frame(normalized_counts), var = "gene_id"),
  by.x = "gene_id",
  by.y = "gene_id"
)
out_norm <- distinct(out_norm)
pca <- prcomp(t(normalized_counts))
gene_variances <- apply(normalized_counts, 1, var)
top_geneids <- names(sort(gene_variances, decreasing = TRUE))[1:20]
top_genes <- tx2gene$gene_name[match(top_geneids, tx2gene$gene_id)]
top_gene_counts <- normalized_counts[top_geneids, ]
rownames(top_gene_counts) <- top_genes
group_label <- if ("group" %in% names(samplesheet)) samplesheet$group else "All"
```

# Interactive figures

Use the PCA and heatmap panels below to inspect sample relationships and the highest-variance genes after normalization.
::: {.panel-tabset}

## 3D PCA

```{r plotly-3d-pca, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
explained_variance <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 2)
explained_variance_text <- paste0("PC", 1:3, " (", explained_variance[1:3], "%)")

pca_data <- as.data.frame(pca$x[, 1:3])
pca_data$sample <- rownames(pca_data)
pca_data$group <- group_label

plot_ly(
  data = pca_data,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~group,
  colors = "Dark2",
  text = ~ paste("Sample:", sample),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 9)
) %>%
  layout(
    title = "3D PCA Plot of Samples",
    scene = list(
      xaxis = list(title = explained_variance_text[1]),
      yaxis = list(title = explained_variance_text[2]),
      zaxis = list(title = explained_variance_text[3])
    ),
    legend = list(title = list(text = "Group"))
  )
```

## 2D PCA

```{r plotly-2d-pca, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
explained_variance <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 2)
explained_variance_text <- paste0("PC", 1:2, " (", explained_variance[1:2], "%)")

pca_data <- as.data.frame(pca$x[, 1:2])
pca_data$sample <- rownames(pca_data)
pca_data$group <- group_label

plot_ly(
  data = pca_data,
  x = ~PC1,
  y = ~PC2,
  color = ~group,
  colors = "Dark2",
  text = ~ paste("Sample:", sample),
  type = "scatter",
  mode = "markers",
  marker = list(size = 12)
) %>%
  layout(
    title = "2D PCA Plot of Samples",
    xaxis = list(title = explained_variance_text[1]),
    yaxis = list(title = explained_variance_text[2]),
    legend = list(title = list(text = "Group"))
  )
```

## Heatmap

```{r plotly-heatmap, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
row_dendrogram <- hclust(dist(log10(as.matrix(top_gene_counts) + 1)))
ordered_rows <- rownames(top_gene_counts)[row_dendrogram$order]
top_gene_counts_ordered <- top_gene_counts[ordered_rows, ]

plot_ly(
  x = colnames(top_gene_counts_ordered),
  y = rownames(top_gene_counts_ordered),
  z = log10(as.matrix(top_gene_counts_ordered) + 1),
  type = "heatmap",
  colors = viridis(256),
  colorbar = list(title = "Expression Level (log10)"),
  showscale = TRUE
) %>%
  layout(
    title = "Heatmap of Top 20 Genes with Highest Variance",
    xaxis = list(title = "Samples", tickangle = 45),
    yaxis = list(title = "Genes (clustered)")
  )
```

:::

# Static figures

Static versions of the PCA and heatmap are provided for reports that are exported to non-interactive formats.

::: {.panel-tabset}

## 2D PCA

```{r ggplot-2d-pca, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
explained_variance <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 2)
explained_variance_text <- paste0("PC", 1:2, " (", explained_variance[1:2], "%)")

pca_data <- as.data.frame(pca$x[, 1:2])
pca_data$sample <- rownames(pca_data)
pca_data$group <- samplesheet$group

ggplot(pca_data, aes(x = PC1, y = PC2, color = group, label = sample)) +
  geom_point(size = 3) +
  geom_text_repel(
    size = 3,
    max.overlaps = 10,
    box.padding = 0.3,
    point.padding = 0.2
  ) +
  labs(
    title = "2D PCA Plot of Samples",
    x = explained_variance_text[1],
    y = explained_variance_text[2]
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.title = element_blank(),
    legend.position = "right"
  ) +
  scale_color_brewer(palette = "Dark2")
```

## Heatmap

```{r ggplot-heatmap, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
log_counts <- log10(as.matrix(top_gene_counts) + 1)
row_dendrogram <- hclust(dist(log_counts))
ordered_genes <- rownames(log_counts)[row_dendrogram$order]
log_counts_ordered <- log_counts[ordered_genes, ]
heatmap_data <- as.data.frame(as.table(log_counts_ordered))
colnames(heatmap_data) <- c("Gene", "Sample", "Expression")

ggplot(heatmap_data, aes(x = Sample, y = Gene, fill = Expression)) +
  geom_tile(color = NA) +
  scale_fill_viridis(name = "Expression Level (log10)", option = "viridis") +
  labs(
    title = "Heatmap of Top 20 Genes with Highest Variance",
    x = "Samples",
    y = "Genes (clustered)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )
```

:::

# Export

```{r export, echo=TRUE, results='markup', warning=FALSE, message=TRUE}
wb <- createWorkbook()
addWorksheet(wb, "Samplesheets")
writeData(wb, "Samplesheets", samplesheet)
addWorksheet(wb, "Norm. counts")
writeData(wb, "Norm. counts", out_norm)
saveWorkbook(wb, "DGEA_all_samples_export.xlsx", overwrite = TRUE)
write.csv(out_norm, "DGEA_all_samples_norm_counts.csv", row.names = FALSE)
```

- Download [DGEA_all_samples_export.xlsx](DGEA_all_samples_export.xlsx).
- Download [DGEA_all_samples_norm_counts.csv](DGEA_all_samples_norm_counts.csv).

# Samplesheet

```{r Samplesheet,  echo=FALSE, results='asis', warning=FALSE, message=FALSE}
datatable(
  samplesheet,
  filter = 'top',
  extensions = c("Buttons"),
  options = list(
    autoWidth = FALSE,
    dom = 'Bt',
    searchHighlight = FALSE,
    buttons = c('copy', 'csv', 'excel', 'print'),
    scrollX = TRUE,
    pageLength = -1,
    paging = FALSE,
    columnDefs = list(
      list(className = 'dt-left', targets = "_all")
    )
  ),
  rownames = FALSE
)
```

# References

<div id="refs"></div>

# R session information

```{r session, echo=FALSE, results='markup', warning=FALSE, message=TRUE}
sessionInfo()
```
