---
title: "Simple Comparison for `r params$sample1` and `r params$sample2`"
author: "`r params$authors`"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
    number_sections: false
    embed-resources: false
    theme: lumen
    code_download: true
    code_folding: hide
bibliography: references.bib
params:
  salmon_dir: ""
  samplesheet_path: ""
  samplesheet: null
  organism: ""
  spikein: ""
  application: ""
  name: ""
  authors: ""
  sample1: ""
  sample2: ""
  filetag: ""
---

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)
library(viridis)
library(DT)
library(data.table)
library(openxlsx)
library(ggrepel)
library(tibble)
library(RColorBrewer)
library(pheatmap)

validate_inputs <- function() {
  stopifnot(nzchar(params$salmon_dir))
  stopifnot(nzchar(params$sample1), nzchar(params$sample2))
  if (!nzchar(params$application)) {
    stop("params$application must be set (nfcore_RNAseq or nfcore_3mRNAseq).")
  }
}

load_samplesheet <- function() {
  if (!is.null(params$samplesheet)) {
    return(params$samplesheet)
  }
  if (!nzchar(params$samplesheet_path)) {
    stop("Provide params$samplesheet or params$samplesheet_path.", call. = FALSE)
  }
  read.csv(params$samplesheet_path, stringsAsFactors = FALSE)
}

validate_inputs()
samplesheet <- load_samplesheet()
rownames(samplesheet) <- samplesheet$sample

organism <- params$organism
salmon_dir <- params$salmon_dir
filetag <- if (nzchar(params$filetag)) params$filetag else paste0(params$sample2, "_vs_", params$sample1)
annotation_db <- NULL

if (organism == "mmusculus") {
  library(org.Mm.eg.db)
  annotation_db <- org.Mm.eg.db
} else if (organism == "hsapiens") {
  library(org.Hs.eg.db)
  annotation_db <- org.Hs.eg.db
} else if (organism == "rnorvegicus") {
  library(org.Rn.eg.db)
  annotation_db <- org.Rn.eg.db
} else if (organism == "sscrofa" && requireNamespace("org.Ss.eg.db", quietly = TRUE)) {
  library(org.Ss.eg.db)
  annotation_db <- org.Ss.eg.db
} else {
  warning("No annotation package available for organism: ", organism, ". Gene full-name annotation will be skipped.")
}

counts_file <- if (identical(params$application, "nfcore_RNAseq")) {
  file.path(salmon_dir, "salmon.merged.gene_counts_length_scaled.tsv")
} else {
  file.path(salmon_dir, "salmon.merged.gene_counts_scaled.tsv")
}
if (!file.exists(counts_file)) {
  stop("Missing counts file: ", counts_file, call. = FALSE)
}

ct <- fread(counts_file)
required_cols <- c("gene_id", "gene_name", params$sample1, params$sample2)
missing_cols <- setdiff(required_cols, names(ct))
if (length(missing_cols) > 0) {
  stop("Missing columns in counts file: ", paste(missing_cols, collapse = ", "))
}

ct[[paste0(params$sample1, "_log10")]] <- log10(ct[[params$sample1]] + 1)
ct[[paste0(params$sample2, "_log10")]] <- log10(ct[[params$sample2]] + 1)
```

### [Back to the main page](../Analysis_Report_RNAseq.html)

## Scatter plot

```{r scatter-plot, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
scatter_plot <- plot_ly(
  data = ct,
  x = ~get(paste0(params$sample1, "_log10")),
  y = ~get(paste0(params$sample2, "_log10")),
  text = ~paste("Gene:", gene_name),
  mode = "markers",
  type = "scatter",
  marker = list(size = 5, color = "royalblue", opacity = 0.7)
) %>%
  layout(
    title = paste("Scatter Plot of Gene Expression Between", params$sample1, "and", params$sample2),
    xaxis = list(title = paste(params$sample1, "Expression")),
    yaxis = list(title = paste(params$sample2, "Expression"))
  )

scatter_plot
```

## MA plot

```{r ma-plot, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
sample1_adjusted <- ct[[params$sample1]] + 1
sample2_adjusted <- ct[[params$sample2]] + 1

ct$M <- log2(sample2_adjusted / sample1_adjusted)
ct$A <- (log2(sample2_adjusted) + log2(sample1_adjusted)) / 2

ma_plot <- plot_ly(
  data = ct,
  x = ~A,
  y = ~M,
  text = ~paste("Gene:", gene_name),
  mode = "markers",
  type = "scatter",
  marker = list(size = 5, color = "royalblue", opacity = 0.7)
) %>%
  layout(
    title = paste("MA Plot of Gene Expression Between", params$sample1, "and", params$sample2),
    xaxis = list(title = "Mean Expression (A)"),
    yaxis = list(title = "Log2 Fold Change (M)")
  )

ma_plot
```

## Statistics

```{r statistics, layout="l-body-outset", echo=TRUE, results='asis', warning=FALSE, message=FALSE}
datatable(
  ct,
  extensions = c("FixedColumns", "Buttons"),
  filter = "top",
  options = list(
    autoWidth = FALSE,
    dom = "Blftip",
    pageLength = 10,
    searchHighlight = FALSE,
    buttons = c("copy", "csv", "excel", "print"),
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 2)
  ),
  class = c("compact cell-border stripe hover"),
  rownames = FALSE
) %>%
  formatRound(columns = 3:10, digits = 2)
```

# Export

```{r export, echo=TRUE, results='markup', warning=FALSE, message=TRUE}
if (!is.null(annotation_db)) {
  ct <- ct %>%
    mutate(
      gene_full_name = mapIds(
        annotation_db,
        keys = sub("\\..*", "", gene_id),
        column = "GENENAME",
        keytype = "ENSEMBL",
        multiVals = "first"
      )
    ) %>%
    relocate(gene_full_name, .after = gene_name)
} else {
  ct <- ct %>%
    mutate(gene_full_name = NA_character_) %>%
    relocate(gene_full_name, .after = gene_name)
}

export_filename <- paste0("SimpleComparison_", filetag, "_export.xlsx")
wb <- createWorkbook()
addWorksheet(wb, "Statistics")
writeData(wb, "Statistics", ct, row.names = FALSE)
saveWorkbook(wb, export_filename, overwrite = TRUE)

export_csv <- paste0("SimpleComparison_", filetag, "_stat.csv")
write.csv(ct, export_csv, row.names = FALSE)
```

- Please download the export excel file: [`r export_filename`](`r export_filename`)
- Or download the CSV file: [`r export_csv`](`r export_csv`)
- If you desire to annotate the volcano plot and annotate the genes you prefer, please try out our web app [AnnoFig](https://genomics.rwth-aachen.de/annofig/) with the CSV file above.

# References

<div id="refs"></div>

# R session information

```{r session, echo=FALSE, results='markup', warning=FALSE, message=TRUE}
sessionInfo()
```
