---
title: "00 Study Overview"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    toc-location: left
    html-math-method: katex
    code-fold: true
    code-tools: true
    embed-resources: true
    page-layout: article
    css: report-style.css
    html-table-processing: none
    theme: lumen
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
source("dnam_inputs.R")
source("report_helpers.R")
cfg <- load_project_config("config/datasets.toml")
read_if <- function(path) if (file.exists(path)) readr::read_csv(path, show_col_types = FALSE) else NULL
run_summary <- read_if("results/tables/input_run_summary.csv")
sample_inventory <- read_if("results/tables/sample_inventory_overview.csv")
if (is.null(sample_inventory)) sample_inventory <- read_if("results/tables/sample_inventory.csv")
overview_group_counts <- read_if("results/tables/overview_group_counts.csv")
batch_summary <- read_if("results/tables/batch_correction_summary.csv")
comparison_index <- read_if("results/tables/comparison_index.csv")
overview_tbl <- tibble::tibble(
  metric = c("project_name", "configured_array_type", "configured_genome_build", "n_enabled_datasets", "n_samples", "n_configured_comparisons"),
  value = c(
    project_name,
    coalesce_chr(cfg$project$array_type, "AUTO"),
    coalesce_chr(cfg$project$genome_build, "hg38"),
    if (is.null(run_summary)) NA else nrow(run_summary),
    if (is.null(sample_inventory)) NA else nrow(sample_inventory),
    if (is.null(comparison_index) || "note" %in% names(comparison_index)) 0 else nrow(comparison_index)
  )
)
write_table(overview_tbl, "results/tables/overview_table.csv")
sample_inventory_preview <- if (is.null(sample_inventory)) NULL else preview_report_table(
  sample_inventory,
  max_rows_inline = 50
)
comparison_index_preview <- if (is.null(comparison_index)) NULL else preview_report_table(
  comparison_index,
  max_rows_inline = 30
)
```

## Study Summary

This report summarizes the rendered methylation array study workspace. The Linkar project itself is treated as the study, so the report name follows the project directory rather than a separate template parameter.

```{r overview-table}
render_report_table(overview_tbl, page_length = 10)
```

## Input Datasets

These datasets were enabled and successfully merged for downstream analysis.

```{r input-runs-plot}
if (!is.null(run_summary) && all(c("dataset_id", "n_samples") %in% names(run_summary))) {
  ds_plot <- run_summary |>
    dplyr::group_by(.data$dataset_id) |>
    dplyr::summarise(n_samples = sum(.data$n_samples, na.rm = TRUE), .groups = "drop") |>
    dplyr::arrange(dplyr::desc(.data$n_samples), .data$dataset_id) |>
    dplyr::mutate(dataset_id = factor(.data$dataset_id, levels = rev(unique(.data$dataset_id))))
  ggplot(ds_plot, aes(x = dataset_id, y = n_samples)) +
    geom_col(fill = "#4C78A8", width = 0.75) +
    coord_flip() +
    labs(
      title = "Samples Per Enabled Dataset",
      x = NULL,
      y = "Samples"
    ) +
    theme_pub()
}
```

```{r input-runs}
if (!is.null(run_summary)) {
  render_report_table(run_summary, preferred = c("dataset_id", "array_type", "n_samples", "n_common_probes", "normalization"), page_length = 10)
} else {
  cat("Input dataset summary not found.\n")
}
```

## Sample Inventory

This sample inventory is the cohort currently wired into the overview reports. If no
overview-specific subset was prepared, the full merged inventory is shown instead.

```{r sample-inventory}
if (!is.null(sample_inventory)) {
  render_report_table(sample_inventory_preview, preferred = c("sample_id", "dataset_id", "group", "subgroup", "batch", "analysis_set", "sex", "age"), page_length = 25)
  report_preview_note(sample_inventory, max_rows_inline = 50, noun = "sample rows")
}
```

## Overview Group Counts

```{r overview-group-counts-plot}
if (!is.null(overview_group_counts) && all(c("group", "n_samples") %in% names(overview_group_counts))) {
  plot_df <- overview_group_counts |>
    dplyr::arrange(dplyr::desc(.data$n_samples)) |>
    dplyr::mutate(group = factor(.data$group, levels = rev(.data$group)))
  ggplot(plot_df, aes(x = group, y = n_samples)) +
    geom_col(fill = "#2A9D8F", width = 0.75) +
    coord_flip() +
    labs(
      title = "Overview Cohort Size By Group",
      x = NULL,
      y = "Samples"
    ) +
    theme_pub()
}
```

```{r overview-group-counts}
if (!is.null(overview_group_counts)) {
  render_report_table(overview_group_counts, preferred = c("group", "n_samples"), page_length = 15)
}
```

## Batch Correction

This table records whether batch-effect removal was applied to the merged cohort matrix and which variables were preserved in the correction model.

```{r batch-summary}
if (!is.null(batch_summary)) {
  render_report_table(batch_summary, page_length = 10)
}
```

## Comparison Index

These are the standalone comparison reports configured in `DNAm_constructor.R`.

```{r comparison-index}
if (!is.null(comparison_index)) {
  render_report_table(comparison_index_preview, preferred = c("report_order", "report_file", "comparison_name", "target_group", "base_group", "n_samples"), page_length = 20)
  report_preview_note(comparison_index, max_rows_inline = 30, noun = "comparison rows")
} else {
  cat("No comparisons have been configured yet.\n")
}
```
