---
title: "02 All-Sample Embeddings"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    toc-location: left
    html-math-method: katex
    code-fold: true
    code-tools: true
    embed-resources: true
    page-layout: article
    css: report-style.css
    html-table-processing: none
    theme: lumen
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
source("dnam_inputs.R")
source("report_helpers.R")
suppressPackageStartupMessages(library(plotly))

read_if <- function(path) if (file.exists(path)) readr::read_csv(path, show_col_types = FALSE) else NULL

raw_prefix <- if (file.exists("results/tables/overview_cohort_raw_pca.csv")) "overview_cohort_raw" else "raw_all_samples"
corrected_prefix <- if (file.exists("results/tables/overview_cohort_corrected_pca.csv")) "overview_cohort_corrected" else "corrected_all_samples"

pretty_dataset_label <- function(x) {
  dplyr::recode(
    x,
    gse197094_epic = "GSE197094 EPIC",
    gse216262_epic = "GSE216262 EPIC",
    gse216262_450k = "GSE216262 450K",
    own_batch_epic = "Own batch EPIC",
    .default = gsub("_", " ", x)
  )
}

pretty_import_group_label <- function(x) {
  dplyr::recode(
    x,
    EPIC_1051943 = "EPIC v1 (1051943)",
    EPIC_1052641 = "EPIC v1 (1052641)",
    EPIC_V2_1105209 = "EPIC v2",
    `450K_622399` = "450K",
    .default = gsub("_", " ", x)
  )
}

normalize_chr <- function(x) {
  x <- as.character(x)
  x[x %in% c("", "NA", "NaN", "NULL")] <- NA_character_
  x
}

format_hover_value <- function(x, digits = 3) {
  if (is.null(x) || length(x) == 0 || is.na(x)) return("NA")
  if (is.numeric(x)) return(formatC(x, format = "f", digits = digits))
  as.character(x)
}

prepare_embedding_df <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df <- tibble::as_tibble(df)
  imp_x <- if ("import_group.x" %in% names(df)) normalize_chr(df[["import_group.x"]]) else rep(NA_character_, nrow(df))
  imp_y <- if ("import_group.y" %in% names(df)) normalize_chr(df[["import_group.y"]]) else rep(NA_character_, nrow(df))
  df$import_group <- dplyr::coalesce(imp_x, imp_y)
  df$dataset_label <- pretty_dataset_label(normalize_chr(df$dataset_id))
  df$batch_label <- pretty_dataset_label(normalize_chr(df$batch))
  df$import_group_label <- pretty_import_group_label(normalize_chr(df$import_group))
  df$qc_status <- ifelse(toupper(normalize_chr(df$qc_fail)) == "TRUE", "QC fail", "Retained")
  df$hover_text <- paste0(
    "<b>", normalize_chr(df$sample_id), "</b>",
    "<br>Group: ", normalize_chr(df$group),
    "<br>Batch: ", df$batch_label,
    "<br>Dataset: ", df$dataset_label,
    "<br>Import group: ", df$import_group_label,
    "<br>Sentrix: ", normalize_chr(df$SentrixBarcode), " / ", normalize_chr(df$SentrixPosition),
    "<br>Failed probes: ", ifelse(is.na(df$frac_failed), "NA", sprintf("%.2f%%", 100 * as.numeric(df$frac_failed))),
    "<br>Mean detection p: ", ifelse(is.na(df$mean_detp), "NA", formatC(as.numeric(df$mean_detp), format = "e", digits = 2))
  )
  df
}

embedding_palette <- function(values) {
  vals <- unique(stats::na.omit(as.character(values)))
  if (length(vals) == 0) return(character())
  setNames(grDevices::hcl.colors(length(vals), palette = "Dark 3"), vals)
}

build_embedding_plotly_2d <- function(df, x, y, color_var, title, subtitle = NULL, height = 420) {
  if (is.null(df) || !all(c(x, y, color_var) %in% names(df))) return(NULL)
  palette_vals <- embedding_palette(df[[color_var]])
  widget <- plotly::plot_ly(
    data = df,
    x = stats::as.formula(paste0("~`", x, "`")),
    y = stats::as.formula(paste0("~`", y, "`")),
    type = "scatter",
    mode = "markers",
    color = stats::as.formula(paste0("~`", color_var, "`")),
    colors = unname(palette_vals),
    text = ~hover_text,
    hovertemplate = paste0("%{text}<br>", x, ": %{x:.3f}<br>", y, ": %{y:.3f}<extra></extra>"),
    marker = list(size = 11, opacity = 0.9, line = list(width = 0.5, color = "white"))
  ) |>
    plotly::layout(
      title = list(text = if (is.null(subtitle)) title else paste0(title, "<br><sup>", subtitle, "</sup>"), x = 0),
      showlegend = TRUE,
      legend = list(
        orientation = "h",
        x = 0,
        xanchor = "left",
        y = -0.22,
        yanchor = "top",
        title = list(text = "")
      ),
      xaxis = list(title = x, zeroline = FALSE),
      yaxis = list(title = y, zeroline = FALSE),
      hoverlabel = list(align = "left"),
      autosize = TRUE,
      margin = list(l = 70, r = 20, t = 80, b = 110),
      height = height
    )
  widget <- plotly::config(widget, responsive = TRUE)
  htmlwidgets::onRender(
    widget,
    "function(el, x) {
      el.style.width = '100%';
      el.style.maxWidth = '100%';
      if (el.parentElement) {
        el.parentElement.style.width = '100%';
        el.parentElement.style.maxWidth = '100%';
      }
      var plotRoot = el.querySelector('.js-plotly-plot');
      if (plotRoot) {
        plotRoot.style.width = '100%';
        plotRoot.style.maxWidth = '100%';
        if (window.Plotly && window.Plotly.Plots) {
          window.Plotly.Plots.resize(plotRoot);
        }
      }
    }"
  )
}

build_embedding_plotly_3d <- function(df, x, y, z, color_var, title, subtitle = NULL, height = 460) {
  if (is.null(df) || !all(c(x, y, z, color_var) %in% names(df))) return(NULL)
  palette_vals <- embedding_palette(df[[color_var]])
  widget <- plotly::plot_ly(
    data = df,
    x = stats::as.formula(paste0("~`", x, "`")),
    y = stats::as.formula(paste0("~`", y, "`")),
    z = stats::as.formula(paste0("~`", z, "`")),
    type = "scatter3d",
    mode = "markers",
    color = stats::as.formula(paste0("~`", color_var, "`")),
    colors = unname(palette_vals),
    text = ~hover_text,
    hovertemplate = paste0(
      "%{text}<br>",
      x, ": %{x:.3f}<br>",
      y, ": %{y:.3f}<br>",
      z, ": %{z:.3f}<extra></extra>"
    ),
    marker = list(size = 4.5, opacity = 0.85, line = list(width = 0.4, color = "white"))
  ) |>
    plotly::layout(
      title = list(text = if (is.null(subtitle)) title else paste0(title, "<br><sup>", subtitle, "</sup>"), x = 0),
      showlegend = TRUE,
      legend = list(
        orientation = "h",
        x = 0,
        xanchor = "left",
        y = -0.18,
        yanchor = "top",
        title = list(text = "")
      ),
      scene = list(
        xaxis = list(title = x),
        yaxis = list(title = y),
        zaxis = list(title = z),
        camera = list(eye = list(x = 1.6, y = 1.6, z = 1.2))
      ),
      margin = list(l = 0, r = 0, t = 80, b = 90),
      height = height
    )
  widget <- plotly::config(widget, responsive = TRUE)
  htmlwidgets::onRender(
    widget,
    "function(el, x) {
      el.style.width = '100%';
      el.style.maxWidth = '100%';
      if (el.parentElement) {
        el.parentElement.style.width = '100%';
        el.parentElement.style.maxWidth = '100%';
      }
      var plotRoot = el.querySelector('.js-plotly-plot');
      if (plotRoot) {
        plotRoot.style.width = '100%';
        plotRoot.style.maxWidth = '100%';
        if (window.Plotly && window.Plotly.Plots) {
          window.Plotly.Plots.resize(plotRoot);
        }
      }
    }"
  )
}

categorical_effect_size <- function(coord, values) {
  coord <- suppressWarnings(as.numeric(coord))
  values <- normalize_chr(values)
  keep <- !is.na(coord) & !is.na(values)
  coord <- coord[keep]
  values <- factor(values[keep])
  if (length(coord) < 4 || nlevels(values) < 2 || nlevels(values) > 24) return(NA_real_)
  grand_mean <- mean(coord)
  ss_total <- sum((coord - grand_mean)^2)
  if (ss_total <= 0) return(0)
  grp_means <- tapply(coord, values, mean)
  grp_n <- table(values)
  ss_between <- sum(as.numeric(grp_n) * (as.numeric(grp_means) - grand_mean)^2)
  unname(ss_between / ss_total)
}

numeric_effect_size <- function(coord, values) {
  coord <- suppressWarnings(as.numeric(coord))
  values <- suppressWarnings(as.numeric(values))
  keep <- !is.na(coord) & !is.na(values)
  if (sum(keep) < 4) return(NA_real_)
  fit <- stats::lm(coord[keep] ~ values[keep])
  unname(summary(fit)$r.squared)
}

build_driver_table <- function(df) {
  if (is.null(df) || nrow(df) == 0 || !all(c("PC1", "PC2") %in% names(df))) return(NULL)
  categorical_vars <- intersect(
    c("batch_label", "dataset_label", "group", "import_group_label", "analysis_set", "SentrixBarcode", "SentrixPosition", "inferred_array_type", "inferred_array_label", "qc_status"),
    names(df)
  )
  numeric_vars <- intersect(c("frac_failed", "mean_detp", "inferred_n_probes"), names(df))

  rows <- list()
  for (var in categorical_vars) {
    rows[[length(rows) + 1]] <- tibble::tibble(
      variable = var,
      variable_type = "categorical",
      pc1_effect = categorical_effect_size(df$PC1, df[[var]]),
      pc2_effect = categorical_effect_size(df$PC2, df[[var]])
    )
  }
  for (var in numeric_vars) {
    rows[[length(rows) + 1]] <- tibble::tibble(
      variable = var,
      variable_type = "numeric",
      pc1_effect = numeric_effect_size(df$PC1, df[[var]]),
      pc2_effect = numeric_effect_size(df$PC2, df[[var]])
    )
  }

  if (length(rows) == 0) return(NULL)
  dplyr::bind_rows(rows) |>
    dplyr::mutate(
      pc1_effect = round(.data$pc1_effect, 3),
      pc2_effect = round(.data$pc2_effect, 3),
      strongest_effect = round(pmax(.data$pc1_effect, .data$pc2_effect, na.rm = TRUE), 3),
      variable = dplyr::recode(
        .data$variable,
        batch_label = "Batch",
        dataset_label = "Dataset",
        import_group_label = "Import group",
        analysis_set = "Analysis set",
        inferred_array_type = "Inferred array type",
        inferred_array_label = "Inferred array label",
        SentrixBarcode = "Sentrix barcode",
        SentrixPosition = "Sentrix position",
        qc_status = "QC status",
        frac_failed = "Failed-probe fraction",
        mean_detp = "Mean detection p-value",
        inferred_n_probes = "Inferred probe count",
        group = "Group"
      )
    ) |>
    dplyr::arrange(dplyr::desc(.data$strongest_effect), .data$variable)
}

raw_pca <- prepare_embedding_df(read_if(file.path("results/tables", paste0(raw_prefix, "_pca.csv"))))
raw_tsne <- prepare_embedding_df(read_if(file.path("results/tables", paste0(raw_prefix, "_tsne.csv"))))
raw_umap <- prepare_embedding_df(read_if(file.path("results/tables", paste0(raw_prefix, "_umap.csv"))))
corrected_pca <- prepare_embedding_df(read_if(file.path("results/tables", paste0(corrected_prefix, "_pca.csv"))))
corrected_tsne <- prepare_embedding_df(read_if(file.path("results/tables", paste0(corrected_prefix, "_tsne.csv"))))
corrected_umap <- prepare_embedding_df(read_if(file.path("results/tables", paste0(corrected_prefix, "_umap.csv"))))

raw_driver_tbl <- build_driver_table(raw_pca)
corrected_driver_tbl <- build_driver_table(corrected_pca)
```

## Raw Merged Cohort

These embeddings summarize the overview cohort before batch correction. Hovering over each point shows the sample identifier, group, batch, import group, Sentrix position, and basic QC values.

### PCA

```{r raw-pca-group}
build_embedding_plotly_2d(
  raw_pca,
  "PC1",
  "PC2",
  "group",
  "Raw PCA by group",
  subtitle = sprintf("%d samples", nrow(raw_pca))
)
```

```{r raw-pca-batch}
build_embedding_plotly_2d(
  raw_pca,
  "PC1",
  "PC2",
  "batch_label",
  "Raw PCA by batch",
  subtitle = "Use hover to inspect samples that drive the visible separation"
)
```

### t-SNE

```{r raw-tsne-group}
build_embedding_plotly_2d(
  raw_tsne,
  "TSNE1",
  "TSNE2",
  "group",
  "Raw t-SNE by group"
)
```

```{r raw-tsne-batch}
build_embedding_plotly_2d(
  raw_tsne,
  "TSNE1",
  "TSNE2",
  "batch_label",
  "Raw t-SNE by batch"
)
```

### UMAP

```{r raw-umap-group}
if (!is.null(raw_umap)) {
  build_embedding_plotly_2d(
    raw_umap,
    "UMAP1",
    "UMAP2",
    "group",
    "Raw UMAP by group"
  )
}
```

```{r raw-umap-batch}
if (!is.null(raw_umap)) {
  build_embedding_plotly_2d(
    raw_umap,
    "UMAP1",
    "UMAP2",
    "batch_label",
    "Raw UMAP by batch"
  )
}
```

### 3D PCA and 3D t-SNE

```{r raw-pca3d}
build_embedding_plotly_3d(
  raw_pca,
  "PC1",
  "PC2",
  "PC3",
  "group",
  "Raw 3D PCA by group",
  subtitle = sprintf("%d samples", nrow(raw_pca))
)
```

```{r raw-tsne3d}
build_embedding_plotly_3d(
  raw_tsne,
  "TSNE1",
  "TSNE2",
  "TSNE3",
  "group",
  "Raw 3D t-SNE by group"
)
```

## Potential Drivers Of The Separation

When the raw PCA shows strong clustering by batch, the next question is whether that separation is most aligned with:

- technical batch labels
- dataset of origin
- array import group or inferred array type
- Sentrix barcode or position
- QC metrics such as failed-probe fraction or mean detection p-value
- biological group itself

The table below scores how strongly each available variable aligns with `PC1` and `PC2` in the raw PCA. For categorical variables this is a simple variance-explained estimate; for numeric variables it is an `R²` from a one-variable linear model. Higher values suggest that variable is a stronger candidate driver of the separation.

```{r raw-driver-table}
if (!is.null(raw_driver_tbl)) {
  render_report_table(
    preview_report_table(raw_driver_tbl, max_rows_inline = 12),
    preferred = c("variable", "variable_type", "pc1_effect", "pc2_effect", "strongest_effect"),
    page_length = 12
  )
  report_preview_note(raw_driver_tbl, max_rows_inline = 12, noun = "driver rows")
}
```

## Batch-Corrected Cohort

When batch correction is enabled, the corrected embeddings below show whether technical separation has been reduced while preserving biological structure.

### PCA

```{r corrected-pca-group}
if (!is.null(corrected_pca)) {
  build_embedding_plotly_2d(
    corrected_pca,
    "PC1",
    "PC2",
    "group",
    "Corrected PCA by group"
  )
}
```

```{r corrected-pca-batch}
if (!is.null(corrected_pca)) {
  build_embedding_plotly_2d(
    corrected_pca,
    "PC1",
    "PC2",
    "batch_label",
    "Corrected PCA by batch"
  )
}
```

### t-SNE

```{r corrected-tsne-group}
if (!is.null(corrected_tsne)) {
  build_embedding_plotly_2d(
    corrected_tsne,
    "TSNE1",
    "TSNE2",
    "group",
    "Corrected t-SNE by group"
  )
}
```

```{r corrected-tsne-batch}
if (!is.null(corrected_tsne)) {
  build_embedding_plotly_2d(
    corrected_tsne,
    "TSNE1",
    "TSNE2",
    "batch_label",
    "Corrected t-SNE by batch"
  )
}
```

### UMAP

```{r corrected-umap-group}
if (!is.null(corrected_umap)) {
  build_embedding_plotly_2d(
    corrected_umap,
    "UMAP1",
    "UMAP2",
    "group",
    "Corrected UMAP by group"
  )
}
```

```{r corrected-umap-batch}
if (!is.null(corrected_umap)) {
  build_embedding_plotly_2d(
    corrected_umap,
    "UMAP1",
    "UMAP2",
    "batch_label",
    "Corrected UMAP by batch"
  )
}
```

### 3D PCA and 3D t-SNE

```{r corrected-pca3d}
if (!is.null(corrected_pca)) {
  build_embedding_plotly_3d(
    corrected_pca,
    "PC1",
    "PC2",
    "PC3",
    "group",
    "Corrected 3D PCA by group"
  )
}
```

```{r corrected-tsne3d}
if (!is.null(corrected_tsne)) {
  build_embedding_plotly_3d(
    corrected_tsne,
    "TSNE1",
    "TSNE2",
    "TSNE3",
    "group",
    "Corrected 3D t-SNE by group"
  )
}
```

## What Else To Check

Beyond the embedding plots themselves, the most useful follow-up checks for a suspected batch effect are:

- whether `SentrixBarcode` or `SentrixPosition` ranks highly in the raw PCA driver table
- whether `import_group` or `inferred_array_type` explains the split, which would suggest array-version structure rather than wet-lab handling alone
- whether `frac_failed` or `mean_detp` aligns with the separation, which would suggest QC-driven structure
- whether the same separation weakens after correction while the biological groups remain distinguishable

```{r corrected-driver-table}
if (!is.null(corrected_driver_tbl)) {
  render_report_table(
    preview_report_table(corrected_driver_tbl, max_rows_inline = 12),
    preferred = c("variable", "variable_type", "pc1_effect", "pc2_effect", "strongest_effect"),
    page_length = 12
  )
  report_preview_note(corrected_driver_tbl, max_rows_inline = 12, noun = "driver rows")
}
```
