---
title: "03 Batch Diagnostics"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    toc-location: left
    html-math-method: katex
    code-fold: true
    code-tools: true
    embed-resources: true
    page-layout: article
    css: report-style.css
    html-table-processing: none
    theme: lumen
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
source("dnam_inputs.R")
source("report_helpers.R")
suppressPackageStartupMessages(library(plotly))

read_if <- function(path) if (file.exists(path)) readr::read_csv(path, show_col_types = FALSE) else NULL

pretty_dataset_label <- function(x) {
  dplyr::recode(
    x,
    gse197094_epic = "GSE197094 EPIC",
    gse216262_epic = "GSE216262 EPIC",
    gse216262_450k = "GSE216262 450K",
    own_batch_epic = "Own batch EPIC",
    .default = gsub("_", " ", x)
  )
}

pretty_import_group_label <- function(x) {
  dplyr::recode(
    x,
    EPIC_1051943 = "EPIC v1 (1051943)",
    EPIC_1052641 = "EPIC v1 (1052641)",
    EPIC_V2_1105209 = "EPIC v2",
    `450K_622399` = "450K",
    .default = gsub("_", " ", x)
  )
}

normalize_chr <- function(x) {
  x <- as.character(x)
  x[x %in% c("", "NA", "NaN", "NULL")] <- NA_character_
  x
}

prepare_pca_df <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df <- tibble::as_tibble(df)
  imp_x <- if ("import_group.x" %in% names(df)) normalize_chr(df[["import_group.x"]]) else rep(NA_character_, nrow(df))
  imp_y <- if ("import_group.y" %in% names(df)) normalize_chr(df[["import_group.y"]]) else rep(NA_character_, nrow(df))
  df$import_group <- dplyr::coalesce(imp_x, imp_y)
  df$dataset_label <- pretty_dataset_label(normalize_chr(df$dataset_id))
  df$batch_label <- pretty_dataset_label(normalize_chr(df$batch))
  df$import_group_label <- pretty_import_group_label(normalize_chr(df$import_group))
  df$hover_text <- paste0(
    "<b>", normalize_chr(df$sample_id), "</b>",
    "<br>Group: ", normalize_chr(df$group),
    "<br>Batch: ", df$batch_label,
    "<br>Dataset: ", df$dataset_label,
    "<br>Import group: ", df$import_group_label,
    "<br>Sentrix: ", normalize_chr(df$SentrixBarcode), " / ", normalize_chr(df$SentrixPosition),
    "<br>PC1: ", formatC(df$PC1, digits = 3, format = "f"),
    "<br>PC2: ", formatC(df$PC2, digits = 3, format = "f")
  )
  df
}

categorical_effect_size <- function(coord, values) {
  coord <- suppressWarnings(as.numeric(coord))
  values <- normalize_chr(values)
  keep <- !is.na(coord) & !is.na(values)
  coord <- coord[keep]
  values <- factor(values[keep])
  if (length(coord) < 4 || nlevels(values) < 2 || nlevels(values) > 24) return(NA_real_)
  grand_mean <- mean(coord)
  ss_total <- sum((coord - grand_mean)^2)
  if (ss_total <= 0) return(0)
  grp_means <- tapply(coord, values, mean)
  grp_n <- table(values)
  ss_between <- sum(as.numeric(grp_n) * (as.numeric(grp_means) - grand_mean)^2)
  unname(ss_between / ss_total)
}

numeric_effect_size <- function(coord, values) {
  coord <- suppressWarnings(as.numeric(coord))
  values <- suppressWarnings(as.numeric(values))
  keep <- !is.na(coord) & !is.na(values)
  if (sum(keep) < 4) return(NA_real_)
  fit <- stats::lm(coord[keep] ~ values[keep])
  unname(summary(fit)$r.squared)
}

build_driver_table <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  categorical_vars <- intersect(
    c("batch_label", "dataset_label", "group", "import_group_label", "analysis_set", "SentrixBarcode", "SentrixPosition", "inferred_array_type", "inferred_array_label"),
    names(df)
  )
  numeric_vars <- intersect(c("frac_failed", "mean_detp", "inferred_n_probes"), names(df))
  rows <- list()
  for (var in categorical_vars) {
    rows[[length(rows) + 1]] <- tibble::tibble(
      variable = var,
      variable_type = "categorical",
      pc1_effect = categorical_effect_size(df$PC1, df[[var]]),
      pc2_effect = categorical_effect_size(df$PC2, df[[var]])
    )
  }
  for (var in numeric_vars) {
    rows[[length(rows) + 1]] <- tibble::tibble(
      variable = var,
      variable_type = "numeric",
      pc1_effect = numeric_effect_size(df$PC1, df[[var]]),
      pc2_effect = numeric_effect_size(df$PC2, df[[var]])
    )
  }
  if (length(rows) == 0) return(NULL)
  dplyr::bind_rows(rows) |>
    dplyr::mutate(
      strongest_effect = pmax(.data$pc1_effect, .data$pc2_effect, na.rm = TRUE),
      pc1_effect = round(.data$pc1_effect, 3),
      pc2_effect = round(.data$pc2_effect, 3),
      strongest_effect = round(.data$strongest_effect, 3),
      variable = dplyr::recode(
        .data$variable,
        batch_label = "Batch",
        dataset_label = "Dataset",
        import_group_label = "Import group",
        analysis_set = "Analysis set",
        inferred_array_type = "Inferred array type",
        inferred_array_label = "Inferred array label",
        SentrixBarcode = "Sentrix barcode",
        SentrixPosition = "Sentrix position",
        frac_failed = "Failed-probe fraction",
        mean_detp = "Mean detection p-value",
        inferred_n_probes = "Inferred probe count",
        group = "Group"
      )
    ) |>
    dplyr::arrange(dplyr::desc(.data$strongest_effect), .data$variable)
}

build_count_table <- function(df, row_var, col_var) {
  if (is.null(df) || !all(c(row_var, col_var) %in% names(df))) return(NULL)
  df |>
    dplyr::mutate(
      row_value = normalize_chr(.data[[row_var]]),
      col_value = normalize_chr(.data[[col_var]])
    ) |>
    dplyr::filter(!is.na(.data$row_value), !is.na(.data$col_value)) |>
    dplyr::count(.data$row_value, .data$col_value, name = "n_samples") |>
    tidyr::pivot_wider(names_from = .data$col_value, values_from = .data$n_samples, values_fill = 0) |>
    dplyr::rename(!!row_var := .data$row_value)
}

build_pc_boxplot <- function(df, pc_col, title) {
  if (is.null(df) || !all(c(pc_col, "batch_label", "group", "hover_text") %in% names(df))) return(NULL)
  palette_vals <- stats::setNames(grDevices::hcl.colors(length(unique(stats::na.omit(df$group))), "Dark 3"), sort(unique(stats::na.omit(df$group))))
  widget <- plotly::plot_ly(
    data = df,
    x = ~batch_label,
    y = stats::as.formula(paste0("~`", pc_col, "`")),
    color = ~group,
    colors = unname(palette_vals),
    type = "box",
    boxpoints = "all",
    jitter = 0.28,
    pointpos = 0,
    text = ~hover_text,
    hovertemplate = paste0("%{text}<br>", pc_col, ": %{y:.3f}<extra></extra>"),
    marker = list(size = 8, opacity = 0.8),
    line = list(width = 1)
  ) |>
    plotly::layout(
      title = list(text = title, x = 0),
      xaxis = list(title = "", tickangle = -15),
      yaxis = list(title = pc_col),
      legend = list(orientation = "h", y = -0.22, x = 0, title = list(text = "")),
      margin = list(l = 70, r = 20, t = 65, b = 95),
      height = 460,
      hoverlabel = list(align = "left")
    )
  widget <- plotly::config(widget, responsive = TRUE)
  htmlwidgets::onRender(
    widget,
    "function(el, x) {
      el.style.width = '100%';
      el.style.maxWidth = '100%';
      if (el.parentElement) {
        el.parentElement.style.width = '100%';
        el.parentElement.style.maxWidth = '100%';
      }
      var plotRoot = el.querySelector('.js-plotly-plot');
      if (plotRoot) {
        plotRoot.style.width = '100%';
        plotRoot.style.maxWidth = '100%';
        if (window.Plotly && window.Plotly.Plots) {
          window.Plotly.Plots.resize(plotRoot);
        }
      }
    }"
  )
}

raw_pca <- prepare_pca_df(read_if("results/tables/overview_cohort_raw_pca.csv"))
if (is.null(raw_pca)) raw_pca <- prepare_pca_df(read_if("results/tables/raw_all_samples_pca.csv"))
corrected_pca <- prepare_pca_df(read_if("results/tables/overview_cohort_corrected_pca.csv"))
if (is.null(corrected_pca)) corrected_pca <- prepare_pca_df(read_if("results/tables/corrected_all_samples_pca.csv"))

raw_driver_tbl <- build_driver_table(raw_pca)
corrected_driver_tbl <- build_driver_table(corrected_pca)
batch_group_tbl <- build_count_table(raw_pca, "batch_label", "group")
batch_dataset_tbl <- build_count_table(raw_pca, "batch_label", "dataset_label")
batch_import_tbl <- build_count_table(raw_pca, "batch_label", "import_group_label")
```

## Why A Separate Batch Diagnostics Report

This report focuses on the question raised by the embedding plots: which technical or metadata variables best explain the visible separation in the raw cohort, and how much of that separation remains after correction.

## Raw PCA Diagnostics

### PC1 By Batch

```{r raw-pc1-batch}
build_pc_boxplot(raw_pca, "PC1", "Raw PC1 by batch")
```

### PC2 By Batch

```{r raw-pc2-batch}
build_pc_boxplot(raw_pca, "PC2", "Raw PC2 by batch")
```

## Candidate Drivers Of The Raw Separation

Higher values below indicate that a variable aligns more strongly with the first two raw principal components.

```{r raw-driver-table}
if (!is.null(raw_driver_tbl)) {
  render_report_table(
    preview_report_table(raw_driver_tbl, max_rows_inline = 15),
    preferred = c("variable", "variable_type", "pc1_effect", "pc2_effect", "strongest_effect")
  )
  report_preview_note(raw_driver_tbl, max_rows_inline = 15, noun = "driver rows")
}
```

## Cross-Tabulations

### Batch By Group

```{r batch-group}
if (!is.null(batch_group_tbl)) {
  render_report_table(batch_group_tbl)
}
```

### Batch By Dataset

```{r batch-dataset}
if (!is.null(batch_dataset_tbl)) {
  render_report_table(batch_dataset_tbl)
}
```

### Batch By Import Group

```{r batch-import}
if (!is.null(batch_import_tbl)) {
  render_report_table(batch_import_tbl)
}
```

## After Batch Correction

### Corrected PC1 By Batch

```{r corrected-pc1-batch}
if (!is.null(corrected_pca)) {
  build_pc_boxplot(corrected_pca, "PC1", "Corrected PC1 by batch")
}
```

### Corrected PC2 By Batch

```{r corrected-pc2-batch}
if (!is.null(corrected_pca)) {
  build_pc_boxplot(corrected_pca, "PC2", "Corrected PC2 by batch")
}
```

### Candidate Drivers After Correction

```{r corrected-driver-table}
if (!is.null(corrected_driver_tbl)) {
  render_report_table(
    preview_report_table(corrected_driver_tbl, max_rows_inline = 15),
    preferred = c("variable", "variable_type", "pc1_effect", "pc2_effect", "strongest_effect")
  )
  report_preview_note(corrected_driver_tbl, max_rows_inline = 15, noun = "driver rows")
}
```

## How To Interpret This

The most useful clues for a true batch effect are:

- `Batch` or `Dataset` ranking highly in the raw driver table
- `Sentrix barcode` or `Sentrix position` ranking highly, which suggests slide/array-placement structure
- `Import group` or `Inferred array type` ranking highly, which suggests EPIC/EPICv2/450K structure
- `Failed-probe fraction` or `Mean detection p-value` ranking highly, which suggests QC-driven separation
- these signals weakening after correction while the biological group structure remains
