---
title: "Comparison Report"
date: today
params:
  bundle_path: ""
  report_title: "Comparison"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    toc-location: left
    html-math-method: katex
    embed-resources: true
    page-layout: article
    css: report-style.css
    html-table-processing: none
    theme: lumen
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
source("dnam_inputs.R")
source("report_helpers.R")
suppressPackageStartupMessages(library(plotly))
bundle <- readRDS(params$bundle_path)
cmp <- bundle$comparison
embeddings <- bundle$embeddings
dmp <- bundle$dmp
dmr <- bundle$dmr
enrichment <- bundle$enrichment
drilldown <- bundle$drilldown

read_if <- function(path) if (!is.null(path) && file.exists(path)) readr::read_csv(path, show_col_types = FALSE) else NULL

pretty_dataset_label <- function(x) {
  dplyr::recode(
    x,
    gse197094_epic = "GSE197094 EPIC",
    gse216262_epic = "GSE216262 EPIC",
    gse216262_450k = "GSE216262 450K",
    own_batch_epic = "Own batch EPIC",
    .default = gsub("_", " ", x)
  )
}

normalize_chr <- function(x) {
  x <- as.character(x)
  x[x %in% c("", "NA", "NaN", "NULL")] <- NA_character_
  x
}

prepare_embedding_df <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df <- tibble::as_tibble(df)
  imp_x <- if ("import_group.x" %in% names(df)) normalize_chr(df[["import_group.x"]]) else rep(NA_character_, nrow(df))
  imp_y <- if ("import_group.y" %in% names(df)) normalize_chr(df[["import_group.y"]]) else rep(NA_character_, nrow(df))
  df$import_group <- dplyr::coalesce(imp_x, imp_y)
  df$batch_label <- pretty_dataset_label(normalize_chr(df$batch))
  df$hover_text <- paste0(
    "<b>", normalize_chr(df$sample_id), "</b>",
    "<br>Group: ", normalize_chr(df$group),
    "<br>Batch: ", df$batch_label,
    "<br>Dataset: ", pretty_dataset_label(normalize_chr(df$dataset_id)),
    "<br>Import group: ", normalize_chr(df$import_group),
    "<br>Sentrix: ", normalize_chr(df$SentrixBarcode), " / ", normalize_chr(df$SentrixPosition)
  )
  df
}

embedding_palette <- function(values) {
  vals <- unique(stats::na.omit(as.character(values)))
  if (length(vals) == 0) return(character())
  setNames(grDevices::hcl.colors(length(vals), palette = "Dark 3"), vals)
}

build_embedding_plotly_2d <- function(df, x, y, color_var, title, height = 380) {
  if (is.null(df) || !all(c(x, y, color_var) %in% names(df))) return(NULL)
  palette_vals <- embedding_palette(df[[color_var]])
  widget <- plotly::plot_ly(
    data = df,
    x = stats::as.formula(paste0("~`", x, "`")),
    y = stats::as.formula(paste0("~`", y, "`")),
    type = "scatter",
    mode = "markers",
    color = stats::as.formula(paste0("~`", color_var, "`")),
    colors = unname(palette_vals),
    text = ~hover_text,
    hovertemplate = paste0("%{text}<br>", x, ": %{x:.3f}<br>", y, ": %{y:.3f}<extra></extra>"),
    marker = list(size = 10, opacity = 0.9, line = list(width = 0.5, color = "white"))
  ) |>
    plotly::layout(
      title = list(text = title, x = 0),
      showlegend = FALSE,
      xaxis = list(title = x, zeroline = FALSE),
      yaxis = list(title = y, zeroline = FALSE),
      hoverlabel = list(align = "left"),
      autosize = TRUE,
      margin = list(l = 60, r = 20, t = 65, b = 50),
      height = height
    )
  widget <- plotly::config(widget, responsive = TRUE)
  htmlwidgets::onRender(
    widget,
    "function(el, x) {
      el.style.width = '100%';
      el.style.maxWidth = '100%';
      if (el.parentElement) {
        el.parentElement.style.width = '100%';
        el.parentElement.style.maxWidth = '100%';
      }
      var plotRoot = el.querySelector('.js-plotly-plot');
      if (plotRoot) {
        plotRoot.style.width = '100%';
        plotRoot.style.maxWidth = '100%';
        if (window.Plotly && window.Plotly.Plots) {
          window.Plotly.Plots.resize(plotRoot);
        }
      }
    }"
  )
}

pca_df <- prepare_embedding_df(read_if(embeddings$artifacts$pca_csv %||% NULL))
tsne_df <- prepare_embedding_df(read_if(embeddings$artifacts$tsne_csv %||% NULL))
umap_df <- prepare_embedding_df(read_if(embeddings$artifacts$umap_csv %||% NULL))
```

## Comparison Definition

This report is a standalone analysis for one comparison defined directly in `DNAm_constructor.R`.

```{r cmp-summary}
cmp_tbl <- tibble::tibble(
  metric = c("comparison_name", "target_group", "base_group", "samples_used", "use_batch_corrected", "covariates", "dmp_alpha", "delta_beta_min", "dmr_enabled", "enrichment_enabled"),
  value = c(
    cmp$name,
    cmp$target_group,
    cmp$base_group,
    nrow(dmp$meta),
    as.character(isTRUE(cmp$use_batch_corrected)),
    ifelse(length(cmp$covariates) > 0, paste(cmp$covariates, collapse = ", "), NA_character_),
    cmp$dmp_alpha,
    cmp$delta_beta_min,
    as.character(isTRUE(cmp$dmr_enabled)),
    as.character(isTRUE(cmp$enrichment_enabled))
  )
)
render_report_table(cmp_tbl, page_length = 10)
```

```{r cmp-design-note}
comparison_groups_present <- sort(unique(as.character(stats::na.omit(dmp$meta$group))))
sample_manifest_full <- NULL
if (file.exists("config/samples.csv")) {
  sample_manifest_full <- suppressWarnings(readr::read_csv("config/samples.csv", show_col_types = FALSE))
}

group_components <- function(final_group_label) {
  if (is.null(sample_manifest_full) || !"sample_id" %in% names(sample_manifest_full)) return(character())
  join_cols <- intersect(c("sample_id", "dataset_id"), names(dmp$meta))
  ref_cols <- intersect(c("sample_id", "dataset_id", "group"), names(sample_manifest_full))
  if (!("group" %in% ref_cols) || length(join_cols) == 0) return(character())
  used_tbl <- dplyr::as_tibble(dmp$meta) |>
    dplyr::select(dplyr::all_of(c(join_cols, "group"))) |>
    dplyr::rename(final_group = .data$group)
  sample_manifest_full |>
    dplyr::select(dplyr::all_of(ref_cols)) |>
    dplyr::inner_join(used_tbl, by = join_cols) |>
    dplyr::filter(.data$final_group == final_group_label) |>
    dplyr::pull(.data$group) |>
    unique() |>
    sort()
}

target_components <- group_components(cmp$target_group)
base_components <- group_components(cmp$base_group)
is_two_group_contrast <- length(comparison_groups_present) == 2 &&
  all(sort(c(cmp$target_group, cmp$base_group)) %in% comparison_groups_present)
is_direct_pairwise <- is_two_group_contrast &&
  length(target_components) <= 1 &&
  length(base_components) <= 1
is_pooled_two_group <- is_two_group_contrast && !is_direct_pairwise

component_text <- function(label, components) {
  if (length(components) == 0) {
    sprintf("<code>%s</code>: composition could not be resolved from the current sample sheet.", label)
  } else if (length(components) == 1) {
    sprintf("<code>%s</code>: direct group <code>%s</code>.", label, components[[1]])
  } else {
    sprintf("<code>%s</code>: pooled/composite group containing %s.", label, paste(sprintf("<code>%s</code>", components), collapse = ", "))
  }
}

if (is_direct_pairwise) {
  htmltools::tagList(
    htmltools::tags$p(
      style = "margin-top: 0.4rem; color: #4A4A4A;",
      htmltools::HTML(sprintf(
        "<strong>Comparison design:</strong> Direct pairwise two-group comparison between <code>%s</code> and <code>%s</code>.",
        cmp$target_group,
        cmp$base_group
      ))
    ),
    htmltools::tags$ul(
      htmltools::tags$li(htmltools::HTML(component_text(cmp$target_group, target_components))),
      htmltools::tags$li(htmltools::HTML(component_text(cmp$base_group, base_components)))
    )
  )
} else if (is_pooled_two_group) {
  htmltools::tagList(
    htmltools::tags$p(
      style = "margin-top: 0.4rem; color: #4A4A4A;",
      htmltools::HTML(sprintf(
        "<strong>Comparison design:</strong> Two-group contrast after relabeling/pooling. The statistical model compares <code>%s</code> versus <code>%s</code>, but at least one side contains samples pooled from multiple original groups.",
        cmp$target_group,
        cmp$base_group
      ))
    ),
    htmltools::tags$ul(
      htmltools::tags$li(htmltools::HTML(component_text(cmp$target_group, target_components))),
      htmltools::tags$li(htmltools::HTML(component_text(cmp$base_group, base_components)))
    )
  )
} else {
  htmltools::tagList(
    htmltools::tags$p(
      style = "margin-top: 0.4rem; color: #4A4A4A;",
      htmltools::HTML(sprintf(
        "<strong>Comparison design:</strong> Not a simple two-group contrast. The filtered comparison set contains %d modeled groups: %s.",
        length(comparison_groups_present),
        paste(sprintf("<code>%s</code>", comparison_groups_present), collapse = ", ")
      ))
    ),
    htmltools::tags$ul(
      htmltools::tags$li(htmltools::HTML(component_text(cmp$target_group, target_components))),
      htmltools::tags$li(htmltools::HTML(component_text(cmp$base_group, base_components)))
    )
  )
}
```

## Sample Manifest

```{r manifest}
manifest_tbl <- preview_report_table(dmp$meta, max_rows_inline = 120)
render_report_table(manifest_tbl, preferred = c("sample_id", "dataset_id", "group", "subgroup", "batch", "analysis_set", "sex", "age"), page_length = 15)
report_preview_note(dmp$meta, max_rows_inline = 120, noun = "samples")
if (!is.null(dmp$artifacts$manifest_csv)) report_download_link(dmp$artifacts$manifest_csv, label = "Download sample manifest")
```

## Comparison Embeddings

These embeddings are computed only from the samples used in this comparison, so they show the contrast-specific structure rather than the full study cohort.

### PCA

```{r comparison-pca-group}
if (!is.null(pca_df)) {
  build_embedding_plotly_2d(pca_df, "PC1", "PC2", "group", "Comparison PCA by group")
}
```

```{r comparison-pca-batch}
if (!is.null(pca_df)) {
  build_embedding_plotly_2d(pca_df, "PC1", "PC2", "batch_label", "Comparison PCA by batch")
}
```

### t-SNE

```{r comparison-tsne-group}
if (!is.null(tsne_df)) {
  build_embedding_plotly_2d(tsne_df, "TSNE1", "TSNE2", "group", "Comparison t-SNE by group")
}
```

```{r comparison-tsne-batch}
if (!is.null(tsne_df)) {
  build_embedding_plotly_2d(tsne_df, "TSNE1", "TSNE2", "batch_label", "Comparison t-SNE by batch")
}
```

### UMAP

```{r comparison-umap-group}
if (!is.null(umap_df)) {
  build_embedding_plotly_2d(umap_df, "UMAP1", "UMAP2", "group", "Comparison UMAP by group")
}
```

## Differential Methylation

```{r dmp-summary}
render_report_table(dmp$summary, preferred = c("comparison_name", "target_group", "base_group", "n_target", "n_base", "n_sig_cpg", "n_target_hyper", "n_target_hypo"), page_length = 10)
```

```{r dmp-volcano}
if (file.exists(dmp$artifacts$volcano_png)) {
  report_image(dmp$artifacts$volcano_png, alt = "DMP volcano plot")
}
```

### Significant CpGs

`direction` indicates methylation change in the target group relative to the base group. `hyper` means higher methylation in the target group, and `hypo` means lower methylation in the target group.

```{r dmp-significant-table}
dmp_sig_tbl <- dmp$table |>
  dplyr::filter(.data$significant) |>
  dplyr::arrange(.data$adj.P.Val, dplyr::desc(abs(.data$delta_beta)))
if (nrow(dmp_sig_tbl) > 0) {
  render_report_table(dmp_sig_tbl, preferred = c("CpG", "delta_beta", "adj.P.Val", "logFC", "direction", "UCSC_RefGene_Name", "Relation_to_Island"), page_length = 15, wide = TRUE)
} else {
  htmltools::tags$p("No CpGs passed the current significance thresholds for this comparison.")
}
```

### Top Ranked CpGs

```{r dmp-table-setup, include=FALSE}
dmp_ranked_tbl <- dmp$table |>
  dplyr::arrange(.data$adj.P.Val, dplyr::desc(abs(.data$delta_beta)))

rank_hyper_tbl <- dmp_ranked_tbl |>
  dplyr::filter(.data$significant, .data$direction == "target_hypermethylated")
rank_hypo_tbl <- dmp_ranked_tbl |>
  dplyr::filter(.data$significant, .data$direction == "target_hypomethylated")
```

#### Top Ranked Target-Hyper CpGs

```{r dmp-table-hyper}
if (nrow(rank_hyper_tbl) > 0) {
  render_report_table(
    rank_hyper_tbl,
    preferred = c("CpG", "delta_beta", "adj.P.Val", "logFC", "UCSC_RefGene_Name", "Relation_to_Island"),
    page_length = 15,
    wide = TRUE
  )
} else {
  htmltools::tags$p("No significant target-hyper CpGs for this comparison.")
}
```

#### Top Ranked Target-Hypo CpGs

```{r dmp-table-hypo}
if (nrow(rank_hypo_tbl) > 0) {
  render_report_table(
    rank_hypo_tbl,
    preferred = c("CpG", "delta_beta", "adj.P.Val", "logFC", "UCSC_RefGene_Name", "Relation_to_Island"),
    page_length = 15,
    wide = TRUE
  )
} else {
  htmltools::tags$p("No significant target-hypo CpGs for this comparison.")
}
```

```{r dmp-table-downloads}
htmltools::tagList(
  if (!is.null(dmp$artifacts$dmp_csv)) report_download_link(dmp$artifacts$dmp_csv, label = "Download full DMP table"),
  if (!is.null(dmp$artifacts$dmp_sig_csv)) report_download_link(dmp$artifacts$dmp_sig_csv, label = "Download significant DMPs")
)
```

## Differential Regions

```{r dmr}
if (is.null(dmr)) {
  htmltools::tags$p("DMR analysis was disabled or did not produce results.")
} else {
  dmr_plot <- NULL
  if (!is.null(dmr$artifacts$dmr_png) && file.exists(dmr$artifacts$dmr_png)) {
    dmr_plot <- report_image(dmr$artifacts$dmr_png, alt = "Ranked DMR plot")
  }
  dmr_tbl <- dmr$table |>
    dplyr::arrange(.data$HMFDR, dplyr::desc(abs(.data$meandiff)))
  dmr_preview <- preview_report_table(dmr_tbl, max_rows_inline = 60)
  htmltools::tagList(
    render_report_table(dmr$summary, page_length = 10),
    dmr_plot,
    render_report_table(dmr_preview, preferred = c("region_id", "seqnames", "start", "end", "no.cpgs", "HMFDR", "meandiff", "direction", "overlapping.genes"), page_length = 15),
    report_preview_note(dmr_tbl, max_rows_inline = 60, noun = "regions"),
    if (!is.null(dmr$artifacts$dmr_csv)) report_download_link(dmr$artifacts$dmr_csv, label = "Download DMR table")
  )
}
```

## Enrichment

```{r enrichment}
if (is.null(enrichment)) {
  htmltools::tags$p("Enrichment analysis was disabled or did not produce results.")
} else {
  enrichment_tbl <- enrichment$table |>
    dplyr::mutate(
      term_name = dplyr::coalesce(.data$Description, .data$TERM),
      ontology = .data$ONTOLOGY
    ) |>
    dplyr::arrange(.data$direction, .data$collection, .data$FDR, .data$P.DE)
  enrichment_preview <- enrichment_tbl |>
    dplyr::group_by(.data$direction, .data$collection) |>
    dplyr::slice_head(n = 10) |>
    dplyr::ungroup()
  htmltools::tagList(
    render_report_table(enrichment_preview, preferred = c("direction", "collection", "ontology", "term_name", "TERM", "N", "DE", "P.DE", "FDR"), page_length = 15, wide = TRUE),
    if (nrow(enrichment_tbl) > nrow(enrichment_preview)) htmltools::tags$p("Showing the top enrichment terms per methylation direction and collection."),
    if (!is.null(enrichment$artifacts$enrichment_csv)) report_download_link(enrichment$artifacts$enrichment_csv, label = "Download enrichment summary")
  )
}
```

## Drilldown

```{r drilldown}
if (is.null(drilldown)) {
  htmltools::tags$p("No drilldown targets were configured for this comparison.")
} else {
  drill_plot <- NULL
  if (!is.null(drilldown$artifacts$heatmap_png) && file.exists(drilldown$artifacts$heatmap_png)) {
    drill_plot <- report_image(drilldown$artifacts$heatmap_png, alt = "Drilldown heatmap")
  }
  drill_tbl <- NULL
  if (!is.null(drilldown$artifacts$summary_csv)) {
    drill_tbl <- readr::read_csv(drilldown$artifacts$summary_csv, show_col_types = FALSE)
  }
  htmltools::tagList(
    drill_plot,
    if (!is.null(drill_tbl)) render_report_table(drill_tbl, page_length = 15),
    if (!is.null(drilldown$artifacts$summary_csv)) report_download_link(drilldown$artifacts$summary_csv, label = "Download drilldown summary"),
    if (!is.null(drilldown$artifacts$beta_csv)) report_download_link(drilldown$artifacts$beta_csv, label = "Download drilldown beta matrix"),
    if (!is.null(drilldown$artifacts$meta_csv)) report_download_link(drilldown$artifacts$meta_csv, label = "Download drilldown sample metadata")
  )
}
```
